An Executable Formal Semantics For A Functional Actor Language

### 2017-12-20 paper/paper.tex
small edits, mostly spacing for 2.3, 2.4

Semantics of actor-based languages have been defined on paper, for example
in~\cite{}. Additionally, there are many implementations of actor-based systems,
including Erlang~\cite{}, Akka~\cite{}, Scheme~\cite{}, and much more. However,
there is currently no formally defined \textit{executable} operational semantics
of an actor-based language. In K, there was previous work done to define the
semantics of Orc, a concurrent programming language~\cite{}, though that work is
incomplete.

@@

\subsection{In-depth Example}
We test our actor semantics on an example of tree recursion~\cite{actor}.
The problem is to calculate the product of all leaves of a binary tree.
The problem is solved by a divide-and-conquer algorithm which calculates the 
results for two subtrees and multiplies the results.
The algorithm is implemented using a \emph{join continuation} as follows.
\begin{verbatim}
let(bind(sink,
  rec(lambda(b, lambda(m, become(b))))),
let(bind(ident,
  lambda(m, m)),
let(bind(joincont,
  rec(lambda(b, lambda(c, lambda(nargs, lambda(firstnum, lambda(num,
    if(eq?(nargs, 0),
       become(app(app(app(b, c), 1), num)),
       seq(become(sink),
           send(c, mult(firstnum, num))))))))))),
let(bind(treeprod,
  rec(lambda(b, lambda(self, lambda(m,
    seq(become(app(b, self)),
        if(isnat(tree(m)),
           send(cust(m), tree(m)),
           letactor(bind(newcust, app(app(app(joincont, cust(m)), 0), nil)),
             seq(send(self, pr(left(tree(m)), newcust)),
                 send(self, pr(right(tree(m)), newcust))))))))))),
let(bind(tr, 
  pr(pr(1, pr(2, pr(3, 4))), pr(pr(5, 6), 7))),
letactor(
  bind(tpactor, app(treeprod, tpactor)),
  bind(custactor, ident),
  send(tpactor, pr(tr, custactor))))))))
\end{verbatim}
Remind that our semantics initializes the initial actor with the behavior of 
executing the above lambda expression.
Therefore after five \texttt{let}-expressions, the initial actor creates two 
actors. One is the \texttt{tpactor} that calculates the tree product.
The other is the \texttt{custactor} that receives the result.
The initial actor sends \texttt{tpactor} a message containing a binary tree 
whose leaves are $1, 2, \dots, 7$, and the address of the customer actor 
\texttt{custactor}.
When the actor system terminates, we get the following configuration
\begin{verbatim}
<T>
  <actors>
    <actor>
      <k> ready ( ... ) </k>
      <id> _93 </id>
    </actor>
    <actor>
      <k> ready ( ... ) </k>
      <id> _179 </id>
    </actor>
    .... ....
    <actor>
      <k> ready ( ... ) </k>
      <id> _45 </id>
    </actor>
    <actor>
      <k> exec ( nil ) </k>
      <id> initactor </id>
    </actor>
    <actor>
      <k> exec ( 5040 ) </k>
      <id> _46 </id>
    </actor>
    ......
    ......
  </actors>
  <messages> .MessageCellBag </messages>
  <definedaddr> ... </definedaddr>
</T>
\end{verbatim}
In the configuration there are many idle actors generated during the 
computation. The customer actor, which in this execution has identity 
\texttt{\_46}, is holding the final product of 5040.

>>>>

Semantics of actor-based languages have been defined on paper, for example
in~\cite{}. Additionally, there are many implementations of actor-based systems,
including Erlang~\cite{}, Akka~\cite{}, Scheme~\cite{}, and much more. However,
there is currently no formally defined \emph{executable} operational semantics
of an actor-based language. In K, there was previous work done to define the
semantics of Orc, a concurrent programming language~\cite{}, though that work is
incomplete.

@@

\subsection{In-depth Example}
We test our actor semantics on many examples. One of the more complicated test
cases is an example of tree recursion~\cite{actor}.  The problem is to calculate
the product of all leaves of a binary tree.  The problem is solved by a
divide-and-conquer algorithm which calculates the results for two subtrees and
multiplies the results.  The algorithm is implemented using a
\emph{join continuation} as follows.
\small
\begin{verbatim}
let(bind(sink,
  rec(lambda(b, lambda(m, become(b))))),
let(bind(ident,
  lambda(m, m)),
let(bind(joincont,
  rec(lambda(b, lambda(c, lambda(nargs, lambda(firstnum, lambda(num,
    if(eq?(nargs, 0),
       become(app(app(app(b, c), 1), num)),
       seq(become(sink),
           send(c, mult(firstnum, num))))))))))),
let(bind(treeprod,
  rec(lambda(b, lambda(self, lambda(m,
    seq(become(app(b, self)),
        if(isnat(tree(m)),
           send(cust(m), tree(m)),
           letactor(bind(newcust, app(app(app(joincont, cust(m)), 0), nil)),
             seq(send(self, pr(left(tree(m)), newcust)),
                 send(self, pr(right(tree(m)), newcust))))))))))),
let(bind(tr, 
  pr(pr(1, pr(2, pr(3, 4))), pr(pr(5, 6), 7))),
letactor(
  bind(tpactor, app(treeprod, tpactor)),
  bind(custactor, ident),
  send(tpactor, pr(tr, custactor))))))))
\end{verbatim}
\normalsize
Recall that our semantics initializes the initial actor with the behavior of
executing the above lambda expression.
Therefore after five \texttt{let}-expressions, the initial actor creates two 
actors. One is the \texttt{tpactor} that calculates the tree product.
The other is the \texttt{custactor} that receives the result.
The initial actor sends \texttt{tpactor} a message containing a binary tree 
whose leaves are $1, 2, \dots, 7$, and the address of the customer actor 
\texttt{custactor}.
When the actor system terminates, we get the following configuration
\begin{verbatim}
<T>
  <actors>
    <actor>
      <k> ready ( ... ) </k>
      <id> _93 </id>
    </actor>
    <actor>
      <k> ready ( ... ) </k>
      <id> _179 </id>
    </actor>
    .... ....
    <actor>
      <k> ready ( ... ) </k>
      <id> _45 </id>
    </actor>
    <actor>
      <k> exec ( nil ) </k>
      <id> initactor </id>
    </actor>
    <actor>
      <k> exec ( 5040 ) </k>
      <id> _46 </id>
    </actor>
    ......
    ......
  </actors>
  <messages> .MessageCellBag </messages>
  <definedaddr> ... </definedaddr>
</T>
\end{verbatim}
In the configuration there are many idle actors generated during the 
computation. The customer actor, which in this execution has identity 
\texttt{\_46}, is holding the final product of 5040.
